---
title: "HABS_HD_7_Analysis"
output: html_document
date: "2025-07-31"
---

# Does Bilingualism Protect against Cognitive Decline? Cross-sectional and Longitudinal Evidence from Ethnically Diverse Older Adults

```{r}
### download required package
library(sjPlot)
library(readxl)
library(readr)
library(lme4)
library(lmerTest) 
library(sjstats)
library(nlme)
library(dplyr)
library(plyr)
library(pwr)
library(olsrr)
library(car)
library(MatchIt)
library(ggplot2)
library(glmmTMB)
library(tableone)
library(sjPlot)
library(aod)
library(ggeffects) 
library(effects)
library(ggplot2)
library(scales)
library(effectsize) 
library(GGally) 
library(lavaan) 
library(tableone)
library(GGally)
library(corrplot)
library(pheatmap)
library(ppcor)
library(broom.mixed)
library(cowplot)
library(ggsignif)
library(emmeans)
library(ggtext)
library(ggpubr)
```
We loaded required R packages (readxl, readr, lme4, lmerTest, sjstats, nlme, dplyr, plyr, pwr, olsrr, car, MatchIt, ggplot2, glmmTMB, tableone, sjPlot, aod, ggeffects, effects, ggplot2, scales, effectsize, GGally, lavaan, tableone, GGally, corrplot) to support data handling, modeling, and visualisation

```{r}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
HABS_HD_cross_sectional<-read.csv("HABS_HD_7_Cross_Sectional_Analysis.csv")
HABS_HD_cross_sectional$Ethnicity <- factor(HABS_HD_cross_sectional$Ethnicity, levels = c("Hispanic","White","Black")) # so referent is first, ie Hispanic)
HABS_HD_cross_sectional$Bilingualism <- factor(HABS_HD_cross_sectional$Bilingualism, levels = c("Monolingual", "Bilingual"))
HABS_HD_cross_sectional$Gender <- factor(HABS_HD_cross_sectional$Gender, levels = c(0, 1), labels = c("Male", "Female"))
contrasts(HABS_HD_cross_sectional$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(HABS_HD_cross_sectional$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)
```
We coded and/or standardised variables (Bilingualism, Ethnicity, Gender) to ensure consistent directionality and levels for analysis.

## 1 Baseline (cross-sectional) analysis

Descriptive data on bilinguals vs. monolinguals across ethnic groups
```{r}
# Define variables again
vars <- c("Age", "Gender","EducationYears", "MMSE_Total","Ethnicity","CognitiveStatus")
table1 <- CreateTableOne(vars = vars, 
                        strata = c("Bilingualism","Ethnicity"), 
                         data = HABS_HD_cross_sectional, 
                         factorVars = c( "Gender","Interview_Language","Ethnicity","CognitiveStatus"),test= TRUE
                         )
#nonnormal_vars <- c("MMSE_Total")

print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```
We summarised baseline characteristics using stratified descriptive tables (Table 1), focusing on group differences across key demographics and cognitive measures

```{r}
###Test the normality of data distribution
shapiro.test(HABS_HD_cross_sectional$Age)
shapiro.test(HABS_HD_cross_sectional$EducationYears )
shapiro.test(HABS_HD_cross_sectional$MMSE_Total)

###Test the normality of data distribution by ethnic group
HABS_HD_Hispanic<-subset(HABS_HD_cross_sectional, Ethnicity == "Hispanic")
shapiro.test(HABS_HD_Hispanic$Age)
shapiro.test(HABS_HD_Hispanic$EducationYears )
shapiro.test(HABS_HD_Hispanic$MMSE_Total)

HABS_HD_Black<-subset(HABS_HD_cross_sectional, Ethnicity == "Black")
shapiro.test(HABS_HD_Black$Age)
shapiro.test(HABS_HD_Black$EducationYears )
shapiro.test(HABS_HD_Black$MMSE_Total)

HABS_HD_White<-subset(HABS_HD_cross_sectional, Ethnicity == "White")
shapiro.test(HABS_HD_White$Age)
shapiro.test(HABS_HD_White$EducationYears )
shapiro.test(HABS_HD_White$MMSE_Total)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_cross_sectional$Age ~ HABS_HD_cross_sectional$Bilingualism)
kruskal.test(HABS_HD_cross_sectional$EducationYears ~ HABS_HD_cross_sectional$Bilingualism)
kruskal.test(HABS_HD_cross_sectional$MMSE_Total ~ HABS_HD_cross_sectional$Bilingualism)
chisq.test(HABS_HD_cross_sectional$Gender,HABS_HD_cross_sectional$Bilingualism)
chisq.test(HABS_HD_cross_sectional$Ethnicity,HABS_HD_cross_sectional$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_Hispanic$Age ~ HABS_HD_Hispanic$Bilingualism)
kruskal.test(HABS_HD_Hispanic$EducationYears ~ HABS_HD_Hispanic$Bilingualism)
kruskal.test(HABS_HD_Hispanic$MMSE_Total ~ HABS_HD_Hispanic$Bilingualism)
chisq.test(HABS_HD_Hispanic$Gender,HABS_HD_Hispanic$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_Black$Age ~ HABS_HD_Black$Bilingualism)
kruskal.test(HABS_HD_Black$EducationYears ~ HABS_HD_Black$Bilingualism)
kruskal.test(HABS_HD_Black$MMSE_Total ~ HABS_HD_Black$Bilingualism)
chisq.test(HABS_HD_Black$Gender,HABS_HD_Black$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_White$Age ~ HABS_HD_White$Bilingualism)
kruskal.test(HABS_HD_White$EducationYears ~ HABS_HD_White$Bilingualism)
kruskal.test(HABS_HD_White$MMSE_Total ~ HABS_HD_White$Bilingualism)
chisq.test(HABS_HD_White$Gender,HABS_HD_White$Bilingualism)
```

### 1.1 Dementia Status as Outcome

```{r}
###Z-scores
HABS_HD_cross_sectional$Age<- scale(HABS_HD_cross_sectional$Age,center = TRUE,scale = TRUE)
HABS_HD_cross_sectional$EducationYears<- scale(HABS_HD_cross_sectional$EducationYears,center = TRUE,scale = TRUE)
HABS_HD_cross_sectional <- filter(HABS_HD_cross_sectional, !is.na(CognitiveStatus)) ### Some participants cannot be labelled according to their criteria 
HABS_HD_cross_sectional$PD <- ifelse(HABS_HD_cross_sectional$CognitiveStatus == "CU", 0, 1)
model_cognitive <- glm(PD ~ Bilingualism * poly(Age,2)  * Gender * Ethnicity + EducationYears, data = HABS_HD_cross_sectional, family = binomial)
car::Anova(model_cognitive)
exp(cbind(OR = coef(model_cognitive), confint(model_cognitive)))
effectsize(model_cognitive)
eta_squared(model_cognitive) 
```

We fit a regression model with PD as the dependent variable. Predictors included Age, Bilingualism, Ethnicity, and Education; interaction terms captured moderation by age, gender, and ethnicity where specified. Overall effects were evaluated using type-II ANOVA (car::Anova), and effect sizes were computed to quantify the magnitude of associations.

Interaction between Bilingualism and Ethnicity is significant, so break down by ethnicity

```{r}
## Analysis in Hispanic
model_H <- glm(PD ~ Bilingualism + poly(Age,2)+ Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Hispanic'), family = binomial)
car::Anova(model_H) 
effectsize(model_H)
exp(cbind(OR = coef(model_H), confint(model_H)))
eta_squared(model_H)
```

Hispanic Bilinguals MORE likely to be CU

```{r}
## Analysis in Black
model_B <- glm(PD ~ Bilingualism + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Black'), family = binomial)
car::Anova(model_B)
effectsize(model_B)  
exp(cbind(OR = coef(model_B), confint(model_B)))
eta_squared(model_B)
``` 
We fit a regression model with PD as the dependent variable in Black people. Predictors included Age, Bilingualism, and Education; interaction terms captured moderation by age and gender where specified.Results show that Black Bilinguals less likely to be CU!

```{r}
## Analysis in White
model_W <- glm(PD ~ Bilingualism + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='White'), family = binomial)
car::Anova(model_W) 
effectsize(model_W)
exp(cbind(OR = coef(model_W), confint(model_W)))
eta_squared(model_W)
```
No bilingualism effect in White

Function for plotting bar graphs
```{r}
fot.fun <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE, id=NULL) {
  library(plyr)
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm) sum(!is.na(x))
    else       length(x)}
  # This does the effectsize. For each group's data frame, return a vector with
  # N, mean, and sd
  if(is.null(id)){
    datac <- ddply(data, groupvars, .drop=.drop,
                   .fun = function(xx, col) {
                     c(N    = length2(xx[[col]], na.rm=na.rm),
                       mean = mean   (xx[[col]], na.rm=na.rm),
                       median = median (xx[[col]], na.rm=na.rm),
                       sum = sum (xx[[col]], na.rm=na.rm),
                       sd   = sd     (xx[[col]], na.rm=na.rm)  )},
                   measurevar  ) }
  else {
    datac <- ddply(data, groupvars, .drop=.drop,
                   .fun = function(xx, col) {
                     c(id = id,
                       N    = length2(xx[[col]], na.rm=na.rm),
                       mean = mean   (xx[[col]], na.rm=na.rm),
                       median = median (xx[[col]], na.rm=na.rm),
                       sum = sum (xx[[col]], na.rm=na.rm),
                       sd   = sd     (xx[[col]], na.rm=na.rm)
                     ) }, measurevar  )
    datac<-datac[,c(3,1,2,4,5,6,7,8)]  }
  # Rename the "mean" column    
  datac <- reshape:::rename(datac, c("mean" = measurevar))
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  return(datac)
}
```

```{r}
## Plot the probability of being PD across groups
HABS_HD_cross_sectional$PD <- ifelse(HABS_HD_cross_sectional$CognitiveStatus == "CI", 1, 0)

plot_data_ethnicity  <- fot.fun(data=HABS_HD_cross_sectional, "PD", c("Bilingualism","Ethnicity"))

# Example significance data frame
sig_data <- data.frame(
  Ethnicity   = c("Hispanic", "Black"),   # only for these groups
  xmin       = c("Monolingual", "Monolingual"),
  xmax        = c("Bilingual", "Bilingual"),
  y_position  = c(0.55, 0.75),            # adjust based on your scale
  annotation  = c("**", "*")
)

plot_data_ethnicity$Ethnicity <- factor(plot_data_ethnicity$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black" ,"White"))

p1<-ggplot(plot_data_ethnicity, aes(x = Bilingualism, y = PD, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = PD - se, ymax = PD + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("Probability of PD") +
  xlab("Bilingualism") +
  ggtitle("Probability of Probable Dementia (PD) by Bilingualism across Ethic Groups") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.8)) + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
      legend.title = element_text(size = 14, family = "serif"),
      legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif"))+
  # Add significance only for Hispanic and Black
 geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,   
    tip_length = 0,
    size = 1,
    textsize = 5
  )

p1

ggsave("Probability_of_PD.png", plot = p1, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

### 1.2 Cognitive Abilities as Outcome

### 1.2.1 EFA for cognitive measures 
```{r}
# Reverse Trails A and B so higher = better
HABS_HD_cross_sectional$Trails_A_Reversed <- -scale(HABS_HD_cross_sectional$Trails_A_Time)
HABS_HD_cross_sectional$Trails_B_Reversed <- -scale(HABS_HD_cross_sectional$Trails_B_Time)
### Select the target cognitive measures  
HABS_data_z = subset(HABS_HD_cross_sectional, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total','Med_ID','Visit_ID'))
HABS_data_z$Med_ID<-as.factor(HABS_data_z$Med_ID)
HABS_data_z$Visit_ID<-as.factor(HABS_data_z$Visit_ID)
HABS_data_z <- HABS_data_z  # make a copy
HABS_data_z[sapply(HABS_data_z, is.numeric)] <- scale(HABS_data_z[sapply(HABS_data_z, is.numeric)])
## create a z-scored dataset for cognitive measures
efa_data = subset(HABS_data_z, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total')) 
```

```{r}
psych::fa.parallel(efa_data,, fa = "pc")
#### Parallel analysis suggests that the number of components =  2 

# Run PCA
pca_result <- psych::principal(efa_data, nfactors = 2, rotate = "oblimin")
# View results
print(pca_result, digits = 2, cutoff = 0.3)
pca_result$Vaccounted
```

```{r}
two_factor_efa_fit = efa(data = efa_data, nfactors=2, missing="fiml", rotation = "oblimin")
summary(two_factor_efa_fit, fit.measures = TRUE, standardized = TRUE)
predicted_scores <- lavPredict(two_factor_efa_fit)
predicted_scores <- as.data.frame(predicted_scores)
colnames(predicted_scores) <- c("Executive_Function", "Episodic_Memory")
HABS_HD_cross_sectional <- cbind(HABS_HD_cross_sectional, predicted_scores)

```

We conducted exploratory factor analysis with 2 factors to identify the latent structure of the cognitive battery and to guide confirmatory modelling


### 1.2.2 Analysis on Cognitive Performance: Executive_Function domain 
```{r}
a1<-lm(Executive_Function ~ Bilingualism * poly(Age,2)* Gender * Ethnicity + EducationYears, data = HABS_HD_cross_sectional)
car::Anova(a1)
eta_squared(a1)
effectsize(a1)
```

We fit a regression model with Executive_Function as the dependent variable.Predictors included Age, Bilingualism, Ethnicity and Education; interaction terms captured moderation by age, gender, and ethnicity where specified.The results indicate significant interactions between Bilingualism and Gender, and Bilingualism and Ethnicity. 

When the 3-way Interaction breakdown by ethnicity

```{r}
###analysis in Hispanic
model_H <- lm(Executive_Function ~ Bilingualism * Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Hispanic'))
car::Anova(model_H) 
eta_squared(model_H)
effectsize(model_H)
```
Bilingual Hispanics outperformed monolingual Hispanics, the main effect of bilingualism interacts with gender, so break it down

```{r}
##analysis in Black
model_B <- lm(Executive_Function ~ Bilingualism * Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Black'))
car::Anova(model_B) 
summary(model_B)
effectsize(model_B)
eta_squared(model_B)
```
The main effect of bilingualism is not significant in Black, but its interaction with Gender is significant. 

```{r}
### Analysis in Black females
model_BF <- lm(Executive_Function ~ Bilingualism + poly(Age,2) +  EducationYears, data = subset(HABS_HD_cross_sectional,   Ethnicity=='Black'& Gender =='Female' ))
car::Anova(model_BF)  
effectsize(model_BF)
eta_squared(model_BF)
```
Bilingual disadvantage in Black females!

```{r}
### Analysis in Black females
model_BM <- lm(Executive_Function ~ Bilingualism + poly(Age,2) +  EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Black'& Gender =='Male'))
car::Anova(model_BM) 
effectsize(model_BM)
eta_squared(model_BM)
```

```{r}
### analysis in White
model_W <- lm(Executive_Function ~ Bilingualism * Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='White'))
car::Anova(model_W)  
effectsize(model_W)
eta_squared(model_W)
```
The main effect of bilingualism is not significant.

```{r}
HABS_HD_cross_sectional_cleaned<-filter(HABS_HD_cross_sectional, !is.na(Executive_Function))
plot_data  <- fot.fun(data=HABS_HD_cross_sectional_cleaned, "Executive_Function", c("Bilingualism","Ethnicity","Gender"))

plot_data$Ethnicity <- factor(plot_data$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
sig_data <- data.frame(
  group1 = c("Monolingual", "Monolingual", "Monolingual"),
  group2 = c("Bilingual", "Bilingual", "Bilingual"),
  Ethnicity = c("Black", "Hispanic", "Hispanic"),
  Gender = c("Female", "Female", "Male"),
  y.position = c(-0.8, -1, -1.2),  # adjust heights to fit your y-scale
  label = c("**", "**", "**")     # significance levels
)
sig_data$Gender <- factor(sig_data$Gender, levels = c("Female", "Male"))
sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))

p2<-ggplot(plot_data, aes(x = Gender, y = Executive_Function, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Executive_Function - se, ymax = Executive_Function + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("Executive Function") +
  xlab("Gender") +
  ggtitle("Executive Function by Bilingualism Across Ethnic and Gender Groups") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 13, family = "serif")
  ) +
  # Add significance
    geom_text(data = sig_data,
            aes(x = Gender, y = y.position, label = label),
            inherit.aes = FALSE,
            size = 5,
            family = "serif")
p2

ggsave("Executive_Function.png", plot = p2, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

### 1.2.3 Analysis on Cognitive Performance: Episodic Memory
```{r}
a2 <- lm(Episodic_Memory ~ Bilingualism * poly(Age,2) * Gender  * Ethnicity + EducationYears, data = HABS_HD_cross_sectional)  
car::Anova(a2)  
eta_squared(a2)
effectsize(a2)
```

Interaction between Bilingualism and Ethnicity is significant, so break down by ethnicity

```{r}
### Analysis in Hispanic
model_H <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Hispanic'))
car::Anova(model_H)
eta_squared(model_H)
effectsize(model_H)
```
Bilingual advantage in Hispanic people!

```{r}
###Analysis in Black
model_B <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Black'))
car::Anova(model_B)  
eta_squared(model_B)
effectsize(model_B)
```
Bilingual disadvantage in Black people!

```{r}
model_W <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='White'))
car::Anova(model_W) 
eta_squared(model_W)
effectsize(model_W)
```
No bilingual effect in White people.

```{r}
HABS_HD_cross_sectional_cleaned<-filter(HABS_HD_cross_sectional, !is.na(Episodic_Memory))
plot_data  <- fot.fun(data=HABS_HD_cross_sectional_cleaned, "Episodic_Memory", c("Bilingualism","Ethnicity"))
plot_data$Ethnicity <- factor(plot_data$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
# Example significance data frame
sig_data <- data.frame(
  Ethnicity   = c("Hispanic", "Black"),   # only for these groups
  xmin       = c("Monolingual", "Monolingual"),
  xmax        = c("Bilingual", "Bilingual"),
  y_position  = c(-0.65, -0.65),            # adjust based on your scale
  annotation  = c("***", "*")
)

sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))

p3<-ggplot(plot_data, aes(x = Bilingualism, y = Episodic_Memory, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Episodic_Memory - se, ymax = Episodic_Memory + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("Episodic Memory") +
  xlab("Bilingualism") +
  ggtitle("Episodic Memory by Bilingualism Across Ethnic Groups") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  #scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
      legend.title = element_text(size = 14, family = "serif" ),
      legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )+
  # Add significance only for Hispanic and Black
 geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,   
    tip_length = 0,
    size = 1,
    textsize = 5
  )

p3

ggsave("Episodic_Memory.png", plot = p3, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

### 1.3 Language domain

### 1.3.1 Language domain: English

To examine whether bilingualism is associated with a disadvantage in the first language (L1), we compared English performance between English monolinguals and bilinguals who identified English as their primary language. 

```{r}
HABS_HD_cross_sectional_English <- subset(HABS_HD_cross_sectional, AMNART_Er !=" ")
###Z-scores, and invert to bigger better
HABS_HD_cross_sectional_English$AMNART <- -scale(HABS_HD_cross_sectional_English$AMNART_Er, center = TRUE, scale = TRUE)
HABS_HD_cross_sectional_English <- HABS_HD_cross_sectional_English %>% filter(Bilingualism2 %in% c("ME", "BE"))
HABS_HD_cross_sectional_English$Bilingualism2 <- factor(HABS_HD_cross_sectional_English$Bilingualism2, levels = c("ME", "BE"))
contrasts(HABS_HD_cross_sectional_English$Bilingualism2) = c(-1,1)  ### (ME:-1; BE: 1)
```

```{r}
model_English <- lm(AMNART ~ Bilingualism2 * poly(Age,2) * Gender  * Ethnicity + EducationYears, data = HABS_HD_cross_sectional_English)
car::Anova(model_English)
eta_squared(model_English)
effectsize(model_English)
```
Interaction between Bilingualism and Ethnicity is significant, so break down by ethnicity 

```{r}
model_H <- lm(AMNART ~ Bilingualism2 + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='Hispanic'))
car::Anova(model_H)
effectsize(model_H)
eta_squared(model_H)
```
Hispanic bilingual-English are worse (have more errors) than Hispanic monolingual-English

```{r}
##Analysis in Black people
model_B <- lm(AMNART ~ Bilingualism2 + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='Black'))
car::Anova(model_B)
effectsize(model_B)
eta_squared(model_B)
```
The main effect of bilingualism is not significant in Black

```{r}
model_W <- lm(AMNART ~ Bilingualism2 + poly(Age,2)  +Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='White'))
car::Anova(model_W)
effectsize(model_W)
eta_squared(model_W)
```
White bilingual-English are better than White monolingual-English

```{r}
HABS_HD_cross_sectional_English<-filter(HABS_HD_cross_sectional_English, !is.na(AMNART))
plot_data  <- fot.fun(data=HABS_HD_cross_sectional_English, "AMNART", c("Bilingualism2","Ethnicity"))

# Ensure Ethnicity column exists and is factor
plot_data$Ethnicity <- factor(plot_data$Ethnicity, levels = c("Hispanic", "Black", "White"))

sig_data <- data.frame(
  Ethnicity   = c("Hispanic", "White"),   # only for these groups
  xmin       = c("ME", "ME"),
  xmax        = c("BE", "BE"),
  y_position  = c(-0.65, 1.1  ),            # adjust based on your scale
  annotation  = c("*", "**")
)

sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))
# Plot
sp1 <- ggplot(plot_data, aes(x = Bilingualism2, y = AMNART, fill = Bilingualism2)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = AMNART - se, ymax = AMNART + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("AMNART (z-scored)") +
  xlab("Bilingualism") +
  ggtitle("English Performance by Bilingualism Across Ethnic Groups") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(-0.7, 1.2)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  ) +
  geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,
    tip_length = 0,
    size = 0.8,
    textsize = 5
  )

sp1

ggsave("English Performance.png", plot = sp1, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

For English, we found a bilingual disadvantage in Hispanic participants, a bilingual advantage in White participants and no bilingual effect in Black participants. 

### 1.3.2 Language domain: Spanish
```{r}
HABS_HD_WAT<-read.csv("HABS_HD_7_WAT.csv")
###Only Hispanic people have the WAT scores, hence there is no Bilingualism: Ethnicity in the model
HABS_HD_WAT$Gender <- factor(HABS_HD_WAT$Gender,levels = c(0, 1),labels = c("Male", "Female"))
HABS_HD_WAT$Bilingualism2 <- factor(HABS_HD_WAT$Bilingualism2,levels =  c("MS", "BS"))
HABS_HD_WAT$WAT<- scale(HABS_HD_WAT$WAT,center = TRUE, scale = TRUE)
```

```{r}
##Analysis based on Monolingual with Spanish as primary language (MS) vs. Bilinguals with Spanish as L1 (BS)
model_spanish<-lm(WAT ~  Bilingualism2 * poly(Age,2)  * Gender + EducationYears, HABS_HD_WAT)
car::Anova(model_spanish) 
effectsize(model_spanish)
eta_squared(model_spanish)
```

Bilingualism related effects are not significant

#### 2 Longitudinal analysis
```{r}
HABS_HD_longitudinal<-read.csv("HABS_HD_7_Longitudinal_Analysis.csv")
HABS_HD_longitudinal$Ethnicity <- factor(HABS_HD_longitudinal$Ethnicity, levels = c("Hispanic","White","Black")) # so referent is first, ie Hispanic
HABS_HD_longitudinal$Bilingualism <- factor(HABS_HD_longitudinal$Bilingualism, levels = c("Monolingual", "Bilingual"))
HABS_HD_longitudinal$Gender <- factor(HABS_HD_longitudinal$Gender, levels = c(0, 1), labels = c("Male", "Female"))

contrasts(HABS_HD_longitudinal$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(HABS_HD_longitudinal$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)

###Z-scores
HABS_HD_longitudinal$A0<- scale(HABS_HD_longitudinal$A0,center = TRUE,scale = TRUE)
HABS_HD_longitudinal$EducationYears<- scale(HABS_HD_longitudinal$EducationYears,center = TRUE,scale = TRUE)
```

```{r}
### Participants' demographic information
# Keep only one row per participant (assuming "Visit_ID" is the participant ID)
unique_participants <- subset(HABS_HD_longitudinal, Visit_ID =="1")

unique_participants$EducationYears<-as.numeric(unique_participants$EducationYears)
# Variables for the table
vars <- c("A0","Gender", "EducationYears", "MMSE_Total","Max_Visit_ID",
          "Ethnicity","Group")

# Create TableOne stratified by Bilingualism
table1 <- CreateTableOne(vars = vars, 
                         strata = "Bilingualism", 
                         data = unique_participants, 
                         factorVars = c("Gender","Ethnicity","Group","Max_Visit_ID"),
                         test = TRUE)

# Print table
print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```
```{r}
###check the normality of distribution
shapiro.test(unique_participants$Age)
shapiro.test(unique_participants$EducationYears )
shapiro.test(unique_participants$MMSE_Total )
```

```{r}
##Test group differences
kruskal.test(unique_participants$Age ~ unique_participants$Bilingualism)
kruskal.test(unique_participants$EducationYears ~ unique_participants$Bilingualism)
kruskal.test(unique_participants$MMSE_Total ~ unique_participants$Bilingualism)
chisq.test(unique_participants$Gender,unique_participants$Bilingualism)
chisq.test(unique_participants$Ethnicity,unique_participants$Bilingualism)
chisq.test(unique_participants$Max_Visit_ID,unique_participants$Bilingualism)
```

### 2.1 Cognitive Stability in Initially Cognitively Unimpaired Participants

In the longitudinal analysis, only participants who were cognitively unimpaired (CU) at their first visit were included. Therefore, the proportion of individuals classified as 'Stable CU' reflects those who remained cognitively unimpaired across subsequent visits, conditional on being CU at baseline.

```{r}
# Get Med_IDs whose Visit_ID == 1 and CognitiveStatus == "CU"
CU_ids <- HABS_HD_longitudinal %>%
  filter(Visit_ID == 1, CognitiveStatus == "CU") %>%
  pull(Med_ID) %>%
  unique()

# Subset the full data based on those IDs
HABS_HD_longitudinal_CU <- HABS_HD_longitudinal %>%
  filter(Med_ID %in% CU_ids)

##PD coded as 1
HABS_HD_longitudinal_CU$PD<- ifelse(HABS_HD_longitudinal_CU$Group == "Stable CU", 0, 1)

HABS_HD_longitudinal_CU <- subset(HABS_HD_longitudinal_CU, Ethnicity !='Black') 
HABS_HD_longitudinal_CU<-droplevels(HABS_HD_longitudinal_CU)

unique_participants2 <- subset(HABS_HD_longitudinal_CU, Visit_ID =="1")

# Variables for the table
vars <- c("PD","Gender")

# Create TableOne stratified by Bilingualism
table1 <- CreateTableOne(vars = vars, 
                         strata = c("Bilingualism", "Gender"),
                         data = unique_participants2, 
                         factorVars = c("PD"),
                         test = TRUE)

# Print table
print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))

contrasts(HABS_HD_longitudinal_CU$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(HABS_HD_longitudinal_CU$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)
contrasts(HABS_HD_longitudinal_CU$Ethnicity) = c(-1,1) ### (Hispanic:-1; White: 1)
```

```{r}
model_stable <- glm(PD ~ Bilingualism * ( poly(Age,2) + Gender + Ethnicity) + EducationYears, data = HABS_HD_longitudinal_CU, family = binomial)
car::Anova(model_stable)
exp(cbind(OR = coef(model_stable), confint(model_stable)))
effectsize(model_stable)
eta_squared(model_stable) 
```

We fit a regression model with StableCU as the dependent variable. Predictors included A0, Bilingualism; interaction terms captured moderation by age, gender, and ethnicity where specified.

Bilingualism:Gender interaction is significant, so breakdown by Gender.

```{r}
###Analysis in females
model_F <- glm(PD ~ Bilingualism + poly(A0, 2) + Ethnicity + EducationYears,  data = subset(HABS_HD_longitudinal_CU, Gender=='Female'), family = binomial)
car::Anova(model_F)
exp(cbind(OR = coef(model_F), confint(model_F)))
effectsize(model_F)
eta_squared(model_F)
```
```{r}
###Analysis in males
model_M <- glm(PD ~ Bilingualism + poly(A0, 2) + Ethnicity + EducationYears,  data = subset(HABS_HD_longitudinal_CU, Gender=='Male'), family = binomial)
car::Anova(model_M)
effectsize(model_M)
eta_squared(model_M)
```
## Plot the probability of conversion to PD acorss visits
```{r}
table(HABS_HD_longitudinal_CU$PD,HABS_HD_longitudinal_CU$Bilingualism,HABS_HD_longitudinal_CU$Gender)

plot_data  <- fot.fun(data=HABS_HD_longitudinal_CU, "PD", c("Bilingualism","Ethnicity","Gender"))

plot_data <- data.frame(
  Gender = rep(c("Female", "Male"), each = 2),
  Bilingualism = factor(rep(c("Monolingual", "Bilingual"), 2),
                        levels = c("Monolingual", "Bilingual")),  # reference level
  PD = c(0.0428, 0.040, 0.0428, 0.0308),  # proportions for bar height
  count = c(50, 47, 50, 36)         # raw counts for labeling
)
plot_data$se <- plot_data$PD * 0.02

# Significance annotation positions
sig_data <- data.frame(
  Gender = c("Female", "Male"),
  x = 1.5,  # middle between Monolingual (1) and Bilingual (2)
  y = c(max(plot_data$PD[plot_data$Gender=="Female"]) + 0.01,
        max(plot_data$PD[plot_data$Gender=="Male"]) + 0.01),
  label = c("**", "*")
)

# Plot
p4 <- ggplot(plot_data, aes(x = Bilingualism, y = PD, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = PD - se, ymax = PD + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  geom_text(aes(y = PD + se + 0.005, label = count),  # counts above error bars
            position = position_dodge(0.9),
            size = 4,
            family = "serif") +
  facet_wrap(~Gender) +
  ylab("PD Conversion Proportion") +
  xlab("Bilingualism") +
  ggtitle("Probability of Conversion to Dementia by Bilingualism Across Gender") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14,family = "serif"),
    legend.text = element_text(size = 14,family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")  
  ) +
  # Significance stars
  geom_text(
    data = sig_data,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 5,
    family = "serif"
  )

p4

ggsave("Probability_of_PD_Conversion2.png", plot = p4, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

Regression analysis predicting longitudinal cognitive stability (StableCU) revealed a significant interaction between bilingualism and gender. Female bilinguals showed higher risk, while male bilinguals showed lower risk.

### 2.2 Cognitive Abilities as Outcome

### 2.2.1 EFA for cognitive measures 

```{r}
# Reverse Trails A and B so higher = better
HABS_HD_longitudinal$Trails_A_Reversed <- -scale(HABS_HD_longitudinal$Trails_A_Time)
HABS_HD_longitudinal$Trails_B_Reversed <- -scale(HABS_HD_longitudinal$Trails_B_Time)

### Select the target cognitive measures  
HABS_data_z2 = subset(HABS_HD_longitudinal, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total','Med_ID','Visit_ID'))
HABS_data_z2$Med_ID<-as.factor(HABS_data_z2$Med_ID)
HABS_data_z2$Visit_ID<-as.factor(HABS_data_z2$Visit_ID)
HABS_data_z2 <- HABS_data_z2  # make a copy
HABS_data_z2[sapply(HABS_data_z2, is.numeric)] <- scale(HABS_data_z2[sapply(HABS_data_z2, is.numeric)])

## create a z-scored dataset for cognitive measures
efa_data2 = subset(HABS_data_z2, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total')) 
```

```{r}
psych::fa.parallel(efa_data2, fa = "pc")
#### Parallel analysis suggests that the number of components =  2 

# Run PCA
pca_result <- psych::principal(efa_data2, nfactors = 2, rotate = "oblimin")
# View results
print(pca_result, digits = 2, cutoff = 0.3)
pca_result$Vaccounted
```
```{r}
two_factor_efa_fit2 = lavaan::efa(data =efa_data2, nfactors=2, rotation='oblimin', missing="fiml")
summary(two_factor_efa_fit2, fit.measures = TRUE, standardized = TRUE)

predicted_scores2 <- lavPredict(two_factor_efa_fit2)
predicted_scores2 <- as.data.frame(predicted_scores2)
colnames(predicted_scores2) <- c("Executive_Function", "Episodic_Memory")

HABS_HD_longitudinal <- cbind(HABS_HD_longitudinal, predicted_scores2)
```
### 2.2.2 Analysis on Cognitive Performance: Executive_Function domain 

```{r}
HABS_HD_longitudinal$CatPractice <- as.factor(HABS_HD_longitudinal$Visit_ID)
a1<-lmer(Executive_Function ~  A0 * dA * Bilingualism * Gender * Ethnicity + EducationYears + CatPractice + (1|Med_ID), data = HABS_HD_longitudinal)
car::Anova(a1)
eta_squared(a1)
test(emtrends(a1, ~ Bilingualism * Ethnicity, var = "dA"))
effectsize(a1)
```

```{r}
HABS_HD_longitudinal$CatPractice <- as.factor(HABS_HD_longitudinal$Visit_ID)
HABS_HD_longitudinal <- HABS_HD_longitudinal %>%
  mutate(A0_group = ifelse(A0 <= median(A0, na.rm = TRUE), "Young", "Old"),
         A0_group = factor(A0_group, levels = c("Young", "Old"))) 

a1<-lmer(Executive_Function ~  A0_group * dA * Bilingualism * Gender * Ethnicity + EducationYears + CatPractice + (1|Med_ID), data = HABS_HD_longitudinal)
car::Anova(a1)
test(emtrends(a1, ~ A0_group * Bilingualism * Ethnicity, var = "dA"))
```
We fit a mixed-effects model with Executive_Function as the outcome, including participant-level random intercepts to account for repeated measurements.Baseline age (A0) and time since baseline (dA) were modelled as fixed effects, with interactions testing whether longitudinal change differed by bilingual status and demographic factors, Practice effect is also controlled in the model. Inference was based on Wald chi-square tests (car::Anova) with follow-up contrasts where relevant.

The 4-way interaction is significant, so breakdown by ethnicity
```{r}
a1_H<-lmer(Executive_Function ~  A0 * dA * Bilingualism + Gender + EducationYears + CatPractice + (1|Med_ID), data = subset(HABS_HD_longitudinal, Ethnicity=='Hispanic'))
car::Anova(a1_H)
eta_squared(a1_H)
effectsize(a1_H)
```

```{r}
a1_W<-lmer(Executive_Function ~  A0 * dA * Bilingualism + Gender + EducationYears + CatPractice + (1|Med_ID), data = subset(HABS_HD_longitudinal, Ethnicity=='White'))
car::Anova(a1_W)
eta_squared(a1_W)
effectsize(a1_W)
```

```{r}
a1_B<-lmer(Executive_Function ~  A0 * dA * Bilingualism+Gender + EducationYears + CatPractice + (1|Med_ID), data = subset(HABS_HD_longitudinal, Ethnicity=='Black'))
car::Anova(a1_B)
eta_squared(a1_B)
effectsize(a1_B)
```

```{r}
plot_spaghetti_collapsed <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, plot.title, plot.subtitle = NULL) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    # raw points
    geom_point(alpha = 0.3, size = 2, color = "darkorange") +
    # raw trajectories per participant
    geom_line(aes(group = Med_ID), linetype = "dotted", alpha = 0.2, color = "#377EB8") +
    # smooth trend line (quadratic)
    geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE, color = "#377EB8", size = 1) +
    labs(
      x = x_label,
      y = y_label,
      title = plot.title,
      subtitle = plot.subtitle
    ) +
    ylim(ymin, ymax) +
    theme_minimal() +
     theme_bw() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
      plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
      legend.position = "top",
      legend.title = element_text(size = 14, family = "serif"),
      legend.text = element_text(size = 14, family = "serif"),
      axis.title.x = element_text(size = 14, family = "serif"),
      axis.title.y = element_text(size = 14, family = "serif"),
      axis.text.x = element_text(size = 13, family = "serif"),  
      axis.text.y = element_text(size = 13, family = "serif"),
      strip.text = element_text(size = 14, family = "serif")
    )
}

p51<- plot_spaghetti_collapsed(
  data = HABS_HD_longitudinal,
  x_var = "Age",
  y_var = "Executive_Function",
  x_label = "Age",
  y_label = "ExecutiveFunction",
  ymin = min(HABS_HD_longitudinal$Executive_Function, na.rm = TRUE),
  ymax = max(HABS_HD_longitudinal$Executive_Function, na.rm = TRUE),
  plot.title = "Executive Function Changes Across Age",
  plot.subtitle = "Dotted lines: individual trajectories; Blue line: quadratic trend"
)
p51
```

```{r}
# 1. Create A0_group and filter dA <= 6
HABS_HD_longitudinal <- HABS_HD_longitudinal %>%
  mutate(A0_group = ifelse(A0 <= median(A0, na.rm = TRUE), "Young", "Old"),
         A0_group = factor(A0_group, levels = c("Young", "Old"))) %>%
  filter(dA <= 6)

# 2. Fit model including A0_group
m2 <- lmer(
  Executive_Function ~ A0_group * dA * Bilingualism * Gender * Ethnicity +
    CatPractice + EducationYears + (1|Med_ID),
  data = HABS_HD_longitudinal
)

# 3. Generate predictions (collapse across gender, include A0_group for prediction)
pred <- emmeans(
  m2,
  ~ dA * Bilingualism | Ethnicity * A0_group,
  at = list(
    dA = seq(min(HABS_HD_longitudinal$dA), max(HABS_HD_longitudinal$dA), length.out = 50),
    CatPractice = "1",
    EducationYears = mean(HABS_HD_longitudinal$EducationYears, na.rm = TRUE),
    Gender = "Female"  # reference level
  )
)

pred_df <- as.data.frame(pred)
pred_df$Ethnicity <- factor(pred_df$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
HABS_HD_longitudinal$Ethnicity <- factor(HABS_HD_longitudinal$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
# 4. Plot
p52<-ggplot() +
  # Raw participant data
  geom_line(data = HABS_HD_longitudinal,
            aes(x = dA, y = Executive_Function, group = Med_ID),
            color = "orange", linetype = "dotted", alpha = 0.4) +
  geom_point(data = HABS_HD_longitudinal,
             aes(x = dA, y = Executive_Function),
             color = "orange", alpha = 0.4, size = 1.2) +
  
  # Model predictions: colored by Bilingualism, linetype = A0_group
  geom_line(data = pred_df,
            aes(x = dA, y = emmean, color = Bilingualism, linetype = A0_group),
            size = 1.2) +
  
  # Confidence ribbons
  geom_ribbon(data = pred_df,
              aes(x = dA, ymin = lower.CL, ymax = upper.CL,
                  fill = Bilingualism, group = interaction(Bilingualism, A0_group)),
              alpha = 0.15, inherit.aes = FALSE) +
  
  # Labels
  labs(
    x = "Time Differences between Visits (dA)",
    y = "Executive Function (adjusted)",
    title = "Executive Function Trajectories by Bilingualism, Ethnicity and Baseline Age",
    subtitle = "Orange dots: individual raw data; Solid lines: model predictions"
  ) +
  
  # Color/fill
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  
  # Facet by Ethnicity
  facet_wrap(~Ethnicity) +
  
  # Theme
  theme_minimal() +
   theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
    plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )

combined_plot <- ggarrange(
  p51, p52,
  ncol = 1, nrow = 2,
  labels = c("A", "B"),      # optional panel labels
  #common.legend = False,      
  legend = "top"              # legend position
)

ggsave(
  "Executive_Function_longitudinal_combined.png",
  plot = combined_plot,
  width = 11,
  height = 16,
  dpi = 800
)
# ggsave("Executive_Function_longitudinal2.png", plot = p51, 
#        width = 11, height = 6, units = "in", 
#        dpi = 800) 
```
### 2.2.2 Analysis on Cognitive Performance: Long-term Memory (Episodic_Memory)

```{r}
a2<-lmer(Episodic_Memory ~  A0 * dA * Bilingualism  * Gender  * Ethnicity + EducationYears + CatPractice + (1|Med_ID), data = HABS_HD_longitudinal)

test(emtrends(a2, ~ Bilingualism , var = "dA"))
car::Anova(a2) 
eta_squared(a2)
effectsize(a2)

```

dA:Bilingualism is significant, indicating bilingual protective effect

Plot for Episodic_Memory Performance
```{r}
plot_spaghetti_collapsed <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, plot.title, plot.subtitle = NULL) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    # raw points
    geom_point(alpha = 0.3, size = 2, color = "darkorange") +
    # raw trajectories per participant
    geom_line(aes(group = Med_ID), linetype = "dotted", alpha = 0.2, color = "#377EB8") +
    # smooth trend line (quadratic)
    geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE, color = "#377EB8", size = 1) +
    labs(
      x = x_label,
      y = y_label,
      title = plot.title,
      subtitle = plot.subtitle
    ) +
    ylim(ymin, ymax) +
    theme_minimal() +
     theme_bw() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
      plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
      legend.position = "top",
      legend.title = element_text(size = 14, family = "serif"),
      legend.text = element_text(size = 14, family = "serif"),
      axis.title.x = element_text(size = 14, family = "serif"),
      axis.title.y = element_text(size = 14, family = "serif"),
      axis.text.x = element_text(size = 13, family = "serif"),  
      axis.text.y = element_text(size = 13, family = "serif"),
      strip.text = element_text(size = 14, family = "serif")
    )
}

p61 <- plot_spaghetti_collapsed(
  data = HABS_HD_longitudinal,
  x_var = "Age",
  y_var = "Episodic_Memory",
  x_label = "Age",
  y_label = "Episodic Memory",
  ymin = min(HABS_HD_longitudinal$Episodic_Memory, na.rm = TRUE),
  ymax = max(HABS_HD_longitudinal$Episodic_Memory, na.rm = TRUE),
  plot.title = "Episodic Memory Changes Across Age",
  plot.subtitle = "Dotted lines: individual trajectories; Blue line: quadratic trend"
)
p61

```



Bilingualism was associated with a slower decline in episodic memory over time.
```{r}
# 1. Fit the model
m1 <- lmer(Episodic_Memory ~ A0 * dA * Ethnicity * Gender * Bilingualism +
             CatPractice + EducationYears + (1|Med_ID),
           data = HABS_HD_longitudinal)

# 2. Filter raw data for dA <= 6
raw_data_filtered <- HABS_HD_longitudinal %>% 
  filter(dA <= 6)

# 3. Get adjusted predictions for dA × Bilingualism, limiting dA to <= 6
pred <- emmeans(m1, ~ dA * Bilingualism,
                at = list(dA = seq(min(raw_data_filtered$dA),
                                   max(raw_data_filtered$dA),
                                   length.out = 50)))
pred_df <- as.data.frame(pred)

# 4. Plot
p62<-ggplot() +
  # Raw participant trajectories
  geom_line(data = raw_data_filtered,
            aes(x = dA, y = Episodic_Memory, group = Med_ID),
            color = "orange", linetype = "dotted", alpha = 0.5) +
  geom_point(data = raw_data_filtered,
             aes(x = dA, y = Episodic_Memory),
             color = "orange", alpha = 0.5, size = 1.5) +
  # Model-predicted lines
  geom_line(data = pred_df,
            aes(x = dA, y = emmean, color = Bilingualism),
            size = 1.2) +
  # Confidence ribbons
  geom_ribbon(data = pred_df,
              aes(x = dA, ymin = lower.CL, ymax = upper.CL, fill = Bilingualism),
              alpha = 0.2, inherit.aes = FALSE) +
  # Labels
  labs(
    x = "Time Differences between Visits (dA)",
    y = "Episodic Memory (adjusted)",
    title = "Episodic Memory Trajectories by Bilingualism",
    subtitle = "Orange dots: individual raw data; Solid lines: model predictions"
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
   theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
    plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )
p62
combined_plot <- ggarrange(
  p61, p62,
  ncol = 1, nrow = 2,
  labels = c("A", "B"),      # optional panel labels
  #common.legend = False,      
  legend = "top"              # legend position
)

ggsave(
  "Episodic_Memory_longitudinal_combined.png",
  plot = combined_plot,
  width = 11,
  height = 16,
  dpi = 800
)

ggsave("Episodic_Memory_longitudinal2.png", plot = p61, 
       width = 10, height = 6, units = "in", 
       dpi = 800) 
```







