---
title: "HABS_HD_7_Analysis"
output: html_document
date: "2025-07-31"
---

# Does Bilingualism Protect against Cognitive Decline? Cross-sectional and Longitudinal Evidence from Ethnically Diverse Older Adults

```{r}
### download required package
library(sjPlot)
library(readxl)
library(readr)
library(lme4)
library(lmerTest) 
library(sjstats)
library(nlme)
library(dplyr)
library(plyr)
library(pwr)
library(olsrr)
library(car)
library(MatchIt)
library(ggplot2)
library(glmmTMB)
library(tableone)
library(sjPlot)
library(aod)
library(ggeffects) 
library(effects)
library(ggplot2)
library(scales)
library(effectsize) 
library(GGally) 
library(lavaan) 
library(tableone)
library(GGally)
library(corrplot)
library(pheatmap)
library(ppcor)
library(broom.mixed)
library(cowplot)
library(ggsignif)
library(emmeans)
library(ggtext)
library(ggpubr)
library(broom)
library(knitr)
library(papaja) 
```
We loaded required R packages (readxl, readr, lme4, lmerTest, sjstats, nlme, dplyr, plyr, pwr, olsrr, car, MatchIt, ggplot2, glmmTMB, tableone, sjPlot, aod, ggeffects, effects, ggplot2, scales, effectsize, GGally, lavaan, tableone, GGally, corrplot) to support data handling, modeling, and visualisation

```{r}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
HABS_HD_cross_sectional<-read.csv("HABS_HD_7_Cross_Sectional_Analysis.csv")
HABS_HD_cross_sectional$Ethnicity <- factor(HABS_HD_cross_sectional$Ethnicity, levels = c("Hispanic","White","Black")) # so referent is first, ie Hispanic)
HABS_HD_cross_sectional$Bilingualism <- factor(HABS_HD_cross_sectional$Bilingualism, levels = c("Monolingual", "Bilingual"))
HABS_HD_cross_sectional$Gender <- factor(HABS_HD_cross_sectional$Gender, levels = c(0, 1), labels = c("Male", "Female"))
contrasts(HABS_HD_cross_sectional$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(HABS_HD_cross_sectional$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)
```
We coded and/or standardised variables (Bilingualism, Ethnicity, Gender) to ensure consistent directionality and levels for analysis.

## 1 Baseline (cross-sectional) analysis

Descriptive data on bilinguals vs. monolinguals across ethnic groups
```{r}
# Define variables again
vars <- c("Age", "Gender","EducationYears", "MMSE_Total","Ethnicity","CognitiveStatus")
table1 <- CreateTableOne(vars = vars, 
                        strata = c("Bilingualism","Ethnicity"), 
                         data = HABS_HD_cross_sectional, 
                         factorVars = c( "Gender","Interview_Language","Ethnicity","CognitiveStatus"),test= TRUE
                         )
#nonnormal_vars <- c("MMSE_Total")

print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```
We summarised baseline characteristics using stratified descriptive tables (Table 1), focusing on group differences across key demographics and cognitive measures

```{r}
###Test the normality of data distribution
shapiro.test(HABS_HD_cross_sectional$Age)
shapiro.test(HABS_HD_cross_sectional$EducationYears )
shapiro.test(HABS_HD_cross_sectional$MMSE_Total)

###Test the normality of data distribution by ethnic group
HABS_HD_Hispanic<-subset(HABS_HD_cross_sectional, Ethnicity == "Hispanic")
shapiro.test(HABS_HD_Hispanic$Age)
shapiro.test(HABS_HD_Hispanic$EducationYears )
shapiro.test(HABS_HD_Hispanic$MMSE_Total)

HABS_HD_Black<-subset(HABS_HD_cross_sectional, Ethnicity == "Black")
shapiro.test(HABS_HD_Black$Age)
shapiro.test(HABS_HD_Black$EducationYears )
shapiro.test(HABS_HD_Black$MMSE_Total)

HABS_HD_White<-subset(HABS_HD_cross_sectional, Ethnicity == "White")
shapiro.test(HABS_HD_White$Age)
shapiro.test(HABS_HD_White$EducationYears )
shapiro.test(HABS_HD_White$MMSE_Total)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_cross_sectional$Age ~ HABS_HD_cross_sectional$Bilingualism)
kruskal.test(HABS_HD_cross_sectional$EducationYears ~ HABS_HD_cross_sectional$Bilingualism)
kruskal.test(HABS_HD_cross_sectional$MMSE_Total ~ HABS_HD_cross_sectional$Bilingualism)
chisq.test(HABS_HD_cross_sectional$Gender,HABS_HD_cross_sectional$Bilingualism)
chisq.test(HABS_HD_cross_sectional$Ethnicity,HABS_HD_cross_sectional$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_Hispanic$Age ~ HABS_HD_Hispanic$Bilingualism)
kruskal.test(HABS_HD_Hispanic$EducationYears ~ HABS_HD_Hispanic$Bilingualism)
kruskal.test(HABS_HD_Hispanic$MMSE_Total ~ HABS_HD_Hispanic$Bilingualism)
chisq.test(HABS_HD_Hispanic$Gender,HABS_HD_Hispanic$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_Black$Age ~ HABS_HD_Black$Bilingualism)
kruskal.test(HABS_HD_Black$EducationYears ~ HABS_HD_Black$Bilingualism)
kruskal.test(HABS_HD_Black$MMSE_Total ~ HABS_HD_Black$Bilingualism)
chisq.test(HABS_HD_Black$Gender,HABS_HD_Black$Bilingualism)
```

```{r}
###Test the group differences among the background variables 
kruskal.test(HABS_HD_White$Age ~ HABS_HD_White$Bilingualism)
kruskal.test(HABS_HD_White$EducationYears ~ HABS_HD_White$Bilingualism)
kruskal.test(HABS_HD_White$MMSE_Total ~ HABS_HD_White$Bilingualism)
chisq.test(HABS_HD_White$Gender,HABS_HD_White$Bilingualism)
```

### 1.1 Dementia Status as Outcome
```{r}
###Z-scores
HABS_HD_cross_sectional$Age<- scale(HABS_HD_cross_sectional$Age,center = TRUE,scale = TRUE)
HABS_HD_cross_sectional$EducationYears<- scale(HABS_HD_cross_sectional$EducationYears,center = TRUE,scale = TRUE)
HABS_HD_cross_sectional <- filter(HABS_HD_cross_sectional, !is.na(CognitiveStatus)) ### Some participants cannot be labelled according to their criteria 
HABS_HD_cross_sectional$PD <- ifelse(HABS_HD_cross_sectional$CognitiveStatus == "CU", 0, 1)
model_cognitive <- glm(PD ~ Bilingualism * Gender * Ethnicity *  poly(Age,2) + EducationYears, data = HABS_HD_cross_sectional, family = binomial)
car::Anova(model_cognitive)
anova_res <- car::Anova(model_cognitive, type = 2)
# Extract p-values for effects involving Bilingualism
# Select only effects involving Bilingualism
rows_biling <- grep("Bilingualism", rownames(anova_res))
pvals <- anova_res[rows_biling, "Pr(>Chisq)"]
# Holm-Bonferroni
p_holm <- p.adjust(pvals, method = "holm")

# Combine for reporting
results <- data.frame(
  Effect = rownames(anova_res)[rows_biling],
  `LR Chisq` = anova_res[rows_biling, "LR Chisq"],
  df = anova_res[rows_biling, "Df"],
  p_raw = round(pvals, 5),
  p_holm = round(p_holm, 5)
)
results

exp(cbind(OR = coef(model_cognitive), confint(model_cognitive)))
effectsize(model_cognitive)
eta_squared(model_cognitive)
```

We fit a regression model with PD as the dependent variable. Predictors included Age, Bilingualism, Ethnicity, and Education; interaction terms captured moderation by age, gender, and ethnicity where specified. Overall effects were evaluated using type-II ANOVA (car::Anova), and effect sizes were computed to quantify the magnitude of associations.

Function for plotting bar graphs
```{r}
fot.fun <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE, id=NULL) {
  library(plyr)
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm) sum(!is.na(x))
    else       length(x)}
  # This does the effectsize. For each group's data frame, return a vector with
  # N, mean, and sd
  if(is.null(id)){
    datac <- ddply(data, groupvars, .drop=.drop,
                   .fun = function(xx, col) {
                     c(N    = length2(xx[[col]], na.rm=na.rm),
                       mean = mean   (xx[[col]], na.rm=na.rm),
                       median = median (xx[[col]], na.rm=na.rm),
                       sum = sum (xx[[col]], na.rm=na.rm),
                       sd   = sd     (xx[[col]], na.rm=na.rm)  )},
                   measurevar  ) }
  else {
    datac <- ddply(data, groupvars, .drop=.drop,
                   .fun = function(xx, col) {
                     c(id = id,
                       N    = length2(xx[[col]], na.rm=na.rm),
                       mean = mean   (xx[[col]], na.rm=na.rm),
                       median = median (xx[[col]], na.rm=na.rm),
                       sum = sum (xx[[col]], na.rm=na.rm),
                       sd   = sd     (xx[[col]], na.rm=na.rm)
                     ) }, measurevar  )
    datac<-datac[,c(3,1,2,4,5,6,7,8)]  }
  # Rename the "mean" column    
  datac <- reshape:::rename(datac, c("mean" = measurevar))
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  return(datac)
}
```

### 1.2 Cognitive Abilities as Outcome

### 1.2.1 EFA for cognitive measures 
```{r}
# Reverse Trails A and B so higher = better
HABS_HD_cross_sectional$Trails_A_Reversed <- -scale(HABS_HD_cross_sectional$Trails_A_Time)
HABS_HD_cross_sectional$Trails_B_Reversed <- -scale(HABS_HD_cross_sectional$Trails_B_Time)
### Select the target cognitive measures  
HABS_data_z = subset(HABS_HD_cross_sectional, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total','Med_ID','Visit_ID'))
HABS_data_z$Med_ID<-as.factor(HABS_data_z$Med_ID)
HABS_data_z$Visit_ID<-as.factor(HABS_data_z$Visit_ID)
HABS_data_z <- HABS_data_z  # make a copy
HABS_data_z[sapply(HABS_data_z, is.numeric)] <- scale(HABS_data_z[sapply(HABS_data_z, is.numeric)])
## create a z-scored dataset for cognitive measures
efa_data = subset(HABS_data_z, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total')) 
```

```{r}
psych::fa.parallel(efa_data,, fa = "pc")
#### Parallel analysis suggests that the number of components =  2 

# Run PCA
pca_result <- psych::principal(efa_data, nfactors = 2, rotate = "oblimin")
# View results
print(pca_result, digits = 2, cutoff = 0.3)
pca_result$Vaccounted
```

```{r}
two_factor_efa_fit = efa(data = efa_data, nfactors=2, missing="fiml", rotation = "oblimin")
summary(two_factor_efa_fit, fit.measures = TRUE, standardized = TRUE)
predicted_scores <- lavPredict(two_factor_efa_fit)
predicted_scores <- as.data.frame(predicted_scores)
colnames(predicted_scores) <- c("Executive_Function", "Episodic_Memory")
HABS_HD_cross_sectional <- cbind(HABS_HD_cross_sectional, predicted_scores)
```

We conducted exploratory factor analysis with 2 factors to identify the latent structure of the cognitive battery and to guide confirmatory modelling

### 1.2.2 Analysis on Cognitive Performance: Executive_Function domain 
```{r}
a1<-lm(Executive_Function ~ Bilingualism * poly(Age,2) * Gender * Ethnicity + EducationYears, data = HABS_HD_cross_sectional)
car::Anova(a1)
anova_res <- car::Anova(a1, type = 2)
# Extract p-values for effects involving Bilingualism
# Select only effects involving Bilingualism
rows_biling <- grep("Bilingualism", rownames(anova_res))
pvals <- anova_res[rows_biling, "Pr(>F)"]
# Holm-Bonferroni
p_holm  <- p.adjust(pvals, method = "holm")
p_bonf  <- p.adjust(pvals, method = "bonferroni")

# Combine for reporting
results <- data.frame(
  Effect = rownames(anova_res)[rows_biling],
  F_value = anova_res[rows_biling, "F value"],
  df = anova_res[rows_biling, "Df"],
  p_raw = round(pvals, 5),
  p_holm = round(p_holm, 5),
  p_bonf = round(p_bonf, 5)
)
results
eta_squared(a1)
effectsize(a1)
summary(a1)
```

```{r}
###Analysis in Hispanic people
a1_H<-lm(Executive_Function ~ Bilingualism + Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity =="Hispanic"))
car::Anova(a1_H)
eta_squared(a1_H)
effectsize(a1_H)
```

```{r}
###Analysis in Black people
a1_B<-lm(Executive_Function ~ Bilingualism + Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity =="Black"))
car::Anova(a1_B)
eta_squared(a1_B)
effectsize(a1_B)
```

```{r}
###Analysis in White people
a1_W<-lm(Executive_Function ~ Bilingualism + Gender + poly(Age,2) + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity =="White"))
car::Anova(a1_W)
eta_squared(a1_W)
effectsize(a1_W)
```


```{r}
HABS_HD_cross_sectional_cleaned<-filter(HABS_HD_cross_sectional, !is.na(Executive_Function))

HABS_HD_cross_sectional_cleaned <- ddply(HABS_HD_cross_sectional_cleaned, "Ethnicity", mutate, EFz = scale(Executive_Function)[,1])

plot_data  <- fot.fun(data=HABS_HD_cross_sectional_cleaned, "EFz", c("Bilingualism","Ethnicity"))

plot_data$Ethnicity <- factor(plot_data$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
# Example significance data frame
sig_data <- data.frame(
  Ethnicity  = c("Hispanic"),
  xmin       = c("Monolingual"),
  xmax       = c("Bilingual"),
  y_position = c(0.5),
  annotation = c("***")
)
sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))

p2<-ggplot(plot_data, aes(x = Bilingualism, y = EFz, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = EFz - se, ymax = EFz + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("Executive Function") +
  xlab("Bilingualism") +
  ggtitle("Executive Function by Bilingualism Across Ethnic Groups") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  #scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
      legend.title = element_text(size = 14, family = "serif" ),
      legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )+
  # Add significance only for Hispanic and Black
 geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,   
    tip_length = 0,
    size = 1,
    textsize = 5
  )
p2
ggsave("Executive_Function2.png", plot = p2, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

### 1.2.3 Analysis on Cognitive Performance: Episodic Memory

```{r}
a2 <- lm(Episodic_Memory ~ Bilingualism * poly(Age,2) * Gender  * Ethnicity + EducationYears, data = HABS_HD_cross_sectional)
car::Anova(a2)
anova_res <- car::Anova(a2, type = 2)
# Extract p-values for effects involving Bilingualism
# Select only effects involving Bilingualism
rows_biling <- grep("Bilingualism", rownames(anova_res))
pvals <- anova_res[rows_biling, "Pr(>F)"]
# Holm-Bonferroni
p_holm  <- p.adjust(pvals, method = "holm")
p_bonf  <- p.adjust(pvals, method = "bonferroni")

# Combine for reporting
results <- data.frame(
  Effect = rownames(anova_res)[rows_biling],
  F_value = anova_res[rows_biling, "F value"],
  df = anova_res[rows_biling, "Df"],
  p_raw = round(pvals, 5),
  p_holm = round(p_holm, 5),
  p_bonf = round(p_bonf, 5)
)
results
```

Interaction between Bilingualism and Ethnicity is significant, so break down by ethnicity

```{r}
### Analysis in Hispanic
model_H <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Hispanic'))
car::Anova(model_H)
eta_squared(model_H)
effectsize(model_H)
```

```{r}
###Analysis in Black
model_B <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='Black'))
car::Anova(model_B)  
eta_squared(model_B)
effectsize(model_B)
```

```{r}
###Analysis in White
model_W <- lm(Episodic_Memory ~ Bilingualism + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional, Ethnicity=='White'))
car::Anova(model_W) 
eta_squared(model_W)
effectsize(model_W)
```

```{r}
HABS_HD_cross_sectional_cleaned<-filter(HABS_HD_cross_sectional, !is.na(Episodic_Memory))
HABS_HD_cross_sectional_cleaned <- ddply(HABS_HD_cross_sectional_cleaned, "Ethnicity", mutate, EMz = scale(Episodic_Memory)[,1])

plot_data  <- fot.fun(data=HABS_HD_cross_sectional_cleaned, "EMz", c("Bilingualism","Ethnicity"))
plot_data$Ethnicity <- factor(plot_data$Ethnicity,
                                        levels = c("Hispanic", "Black", "White"))
# Example significance data frame
sig_data <- data.frame(
  Ethnicity   = c("Hispanic", "Black"),   # only for these groups
  xmin       = c("Monolingual", "Monolingual"),
  xmax        = c("Bilingual", "Bilingual"),
  y_position  = c(0.4, 0.4),            # adjust based on your scale
  annotation  = c("***", "*")
)

sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))

p3<-ggplot(plot_data, aes(x = Bilingualism, y = EMz, fill = Bilingualism)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = EMz - se, ymax = EMz + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("Episodic Memory") +
  xlab("Bilingualism") +
  ggtitle("Episodic Memory by Bilingualism Across Ethnic Groups") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  #scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
      legend.title = element_text(size = 14, family = "serif" ),
      legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )+
  # Add significance only for Hispanic and Black
 geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position+0.1),
    manual = TRUE,
    inherit.aes = FALSE,   
    tip_length = 0,
    size = 1,
    textsize = 5
  )
p3

ggsave("Episodic_Memory.png", plot = p3, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 

combined_plot <- ggarrange(
  p2, p3,
  ncol = 2, nrow = 1,
  labels = c("A", "B"),      # optional panel labels
  #common.legend = False,      
  legend = "top"              # legend position
)

ggsave("EF_EM_Cross-sectional2.png", plot = combined_plot, 
       width = 16, height = 6, units = "in", 
       dpi = 800) 
```

### 1.3 Language domain

### 1.3.1 Language domain: English

To examine whether bilingualism is associated with a disadvantage in the first language (L1), we compared English performance between English monolinguals and bilinguals who identified English as their primary language. 

```{r}
HABS_HD_cross_sectional_English <- subset(HABS_HD_cross_sectional, AMNART_Er !=" ")
###Z-scores, and invert to bigger better
HABS_HD_cross_sectional_English$AMNART <- -scale(HABS_HD_cross_sectional_English$AMNART_Er, center = TRUE, scale = TRUE)
HABS_HD_cross_sectional_English <- HABS_HD_cross_sectional_English %>% filter(Bilingualism2 %in% c("ME", "BE"))
HABS_HD_cross_sectional_English$Bilingualism2 <- factor(HABS_HD_cross_sectional_English$Bilingualism2, levels = c("ME", "BE"))
contrasts(HABS_HD_cross_sectional_English$Bilingualism2) = c(-1,1)  ### (ME:-1; BE: 1)
```

```{r}
model_English <- lm(AMNART ~ Bilingualism2 * Gender  * Ethnicity * poly(Age,2) + EducationYears, data = HABS_HD_cross_sectional_English)
anova_res <- car::Anova(model_English, type = 2)
# Extract p-values for effects involving Bilingualism
# Select only effects involving Bilingualism
rows_biling <- grep("Bilingualism", rownames(anova_res))
pvals <- anova_res[rows_biling, "Pr(>F)"]
# Holm-Bonferroni
p_holm  <- p.adjust(pvals, method = "holm")
p_bonf  <- p.adjust(pvals, method = "bonferroni")

# Combine for reporting
results <- data.frame(
  Effect = rownames(anova_res)[rows_biling],
  F_value = anova_res[rows_biling, "F value"],
  df = anova_res[rows_biling, "Df"],
  p_raw = round(pvals, 5),
  p_holm = round(p_holm, 5),
  p_bonf = round(p_bonf, 5)
)
results

car::Anova(model_English)
eta_squared(model_English)
summary((model_English))
effectsize(model_English)
```

Interaction between Bilingualism and Ethnicity is significant, so break down by ethnicity 

```{r}
##Analysis in Hispanic people
model_H <- lm(AMNART ~ Bilingualism2 + poly(Age,2) + Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='Hispanic'))
car::Anova(model_H)
effectsize(model_H)
eta_squared(model_H)
```

Hispanic bilingual-English are worse (have more errors) than Hispanic monolingual-English

```{r}
##Analysis in Black people
model_B <- lm(AMNART ~ Bilingualism2 + poly(Age,2)  + Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='Black'))
car::Anova(model_B)
effectsize(model_B)
eta_squared(model_B)
```
The main effect of bilingualism is not significant in Black

```{r}
##Analysis in White people
model_W <- lm(AMNART ~ Bilingualism2 + poly(Age,2)  +Gender + EducationYears, data = subset(HABS_HD_cross_sectional_English, Ethnicity=='White'))
car::Anova(model_W)
effectsize(model_W)
eta_squared(model_W)
```

White bilingual-English are better than White monolingual-English

```{r}
HABS_HD_cross_sectional_English<-filter(HABS_HD_cross_sectional_English, !is.na(AMNART))

HABS_HD_cross_sectional_English <- ddply(HABS_HD_cross_sectional_English, "Ethnicity", mutate, AMNARTz = scale(AMNART)[,1])

plot_data  <- fot.fun(data=HABS_HD_cross_sectional_English, "AMNARTz", c("Bilingualism2","Ethnicity"))

# Ensure Ethnicity column exists and is factor
plot_data$Ethnicity <- factor(plot_data$Ethnicity, levels = c("Hispanic", "Black", "White"))

sig_data <- data.frame(
  Ethnicity   = c("Hispanic", "White"),   # only for these groups
  xmin       = c("ME", "ME"),
  xmax        = c("BE", "BE"),
  y_position  = c(0.7, 0.7  ),            # adjust based on your scale
  annotation  = c("*", "**")
)

sig_data$Ethnicity <- factor(sig_data$Ethnicity,
                             levels = c("Hispanic", "Black", "White"))
# Plot
sp1 <- ggplot(plot_data, aes(x = Bilingualism2, y = AMNARTz, fill = Bilingualism2)) +
  geom_col(position = position_dodge(0.9), width = 0.7) +
  geom_errorbar(aes(ymin = AMNARTz - se, ymax = AMNARTz + se),
                width = 0.2,
                position = position_dodge(0.9)) +
  facet_wrap(~Ethnicity) +
  ylab("AMNART (z-scored)") +
  xlab("Bilingualism") +
  ggtitle("English Performance by Bilingualism Across Ethnic Groups") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  ) +
  geom_signif(
    data = sig_data,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,
    tip_length = 0,
    size = 0.8,
    textsize = 5
  )

sp1

ggsave("English Performance2.png", plot = sp1, 
       width = 8, height = 6, units = "in", 
       dpi = 800) 
```

### 1.3.2 Language domain: Spanish
```{r}
HABS_HD_WAT<-read.csv("HABS_HD_7_WAT.csv")
###Only Hispanic people have the WAT scores, hence there is no Bilingualism: Ethnicity in the model
HABS_HD_WAT$Gender <- factor(HABS_HD_WAT$Gender,levels = c(0, 1),labels = c("Male", "Female"))
HABS_HD_WAT$Bilingualism2 <- factor(HABS_HD_WAT$Bilingualism2,levels =  c("MS", "BS"))
HABS_HD_WAT$WAT<- scale(HABS_HD_WAT$WAT,center = TRUE, scale = TRUE)
```

```{r}
##Analysis based on Monolingual with Spanish as primary language (MS) vs. Bilinguals with Spanish as L1 (BS)
model_spanish<-lm(WAT ~  Bilingualism2  * Gender +  poly(Age,2)  + EducationYears, HABS_HD_WAT)
car::Anova(model_spanish) 
effectsize(model_spanish)
eta_squared(model_spanish)
```

Bilingualism related effects are not significant

#### 2 Longitudinal analysis
```{r}
HABS_HD_longitudinal<-read.csv("HABS_HD_7_Longitudinal_Analysis.csv")
HABS_HD_longitudinal$Ethnicity <- factor(HABS_HD_longitudinal$Ethnicity, levels = c("Hispanic","White","Black")) # so referent is first, ie Hispanic
HABS_HD_longitudinal$Bilingualism <- factor(HABS_HD_longitudinal$Bilingualism, levels = c("Monolingual", "Bilingual"))
HABS_HD_longitudinal$Gender <- factor(HABS_HD_longitudinal$Gender, levels = c(0, 1), labels = c("Male", "Female"))

contrasts(HABS_HD_longitudinal$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(HABS_HD_longitudinal$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)

###Z-scores
HABS_HD_longitudinal$A0<- scale(HABS_HD_longitudinal$A0,center = TRUE,scale = TRUE)
HABS_HD_longitudinal$EducationYears<- scale(HABS_HD_longitudinal$EducationYears,center = TRUE,scale = TRUE)
```

```{r}
### Participants' demographic information
# Keep only one row per participant (assuming "Visit_ID" is the participant ID)
unique_participants <- subset(HABS_HD_longitudinal, Visit_ID =="1")

table(unique_participants$Gender)
summary(HABS_HD_longitudinal$Gender)

unique_participants$EducationYears<-as.numeric(unique_participants$EducationYears)
# Variables for the table
vars <- c("A0","Gender", "EducationYears", "MMSE_Total","Group", "Max_Visit_ID")

# Create TableOne stratified by Bilingualism
table1 <- CreateTableOne(vars = vars, 
                         c("Bilingualism","Ethnicity"), 
                         data = unique_participants, 
                         factorVars = c("Gender","Ethnicity","Group","Max_Visit_ID"),
                         test = TRUE)

# Print table
print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))

# Variables for the table
vars <- c("dA")

# Create TableOne stratified by Bilingualism
table2 <- CreateTableOne(vars = vars, 
                         strata = c("Bilingualism", "Ethnicity"),
                         data = HABS_HD_longitudinal, 
                         factorVars = c("Bilingualism", "Ethnicity"),
                         test = TRUE)
# Print table
print(table2, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

```{r}
###check the normality of distribution
shapiro.test(unique_participants$Age)
shapiro.test(unique_participants$EducationYears )
shapiro.test(unique_participants$MMSE_Total )
```

```{r}
##Test group differences
kruskal.test(unique_participants$Age ~ unique_participants$Bilingualism)
kruskal.test(unique_participants$EducationYears ~ unique_participants$Bilingualism)
kruskal.test(unique_participants$MMSE_Total ~ unique_participants$Bilingualism)
chisq.test(unique_participants$Gender,unique_participants$Bilingualism)
chisq.test(unique_participants$Ethnicity,unique_participants$Bilingualism)
chisq.test(unique_participants$Max_Visit_ID,unique_participants$Bilingualism)
```

### 2.1 Cognitive Stability in Initially Cognitively Unimpaired Participants

In the longitudinal analysis, only participants who were cognitively unimpaired (CU) at their first visit were included. Therefore, the proportion of individuals classified as 'Stable CU' reflects those who remained cognitively unimpaired across subsequent visits, conditional on being CU at baseline.

```{r}
# Get Med_IDs whose Visit_ID == 1 and CognitiveStatus == "CU"
CU_ids <- HABS_HD_longitudinal %>%
  filter(Visit_ID == 1, CognitiveStatus == "CU") %>%
  pull(Med_ID) %>%
  unique()

# Subset the full data based on those IDs
HABS_HD_longitudinal_CU <- HABS_HD_longitudinal %>%
  filter(Med_ID %in% CU_ids)

##PD coded as 1
HABS_HD_longitudinal_CU$PD<- ifelse(HABS_HD_longitudinal_CU$Group == "Stable CU", 0, 1)

HABS_HD_longitudinal_CU2 <- subset(HABS_HD_longitudinal_CU, Ethnicity !='Black') 
HABS_HD_longitudinal_CU2<-droplevels(HABS_HD_longitudinal_CU2)

unique_participants2 <- subset(HABS_HD_longitudinal_CU2, Visit_ID =="1")
table(unique_participants2$Max_Visit_ID)

# Variables for the table
vars <- c("PD")

# Create TableOne stratified by Bilingualism
table1 <- CreateTableOne(vars = vars, 
                         strata = c("Bilingualism", "Max_Visit_ID","Gender"),
                         data = unique_participants2, 
                         factorVars = c("PD"),
                         test = TRUE)

# Print table
print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))

contrasts(unique_participants2$Gender) = c(-1,1)  ### (male:-1; female: 1)
contrasts(unique_participants2$Bilingualism) = c(-1,1) ### (monolingual:-1; bilingual: 1)
contrasts(unique_participants2$Ethnicity) = c(-1,1) ### (Hispanic:-1; White: 1)
```

```{r}
model_stable <- glm(PD ~ Bilingualism * (poly(A0,2)+ Gender + Ethnicity)+ EducationYears, data = unique_participants2, family = binomial)
car::Anova(model_stable)
anova_res <- car::Anova(model_stable, type = 2)
# Extract p-values for effects involving Bilingualism
# Select only effects involving Bilingualism
rows_biling <- grep("Bilingualism", rownames(anova_res))
pvals <- anova_res[rows_biling, "Pr(>Chisq)"]
# Holm-Bonferroni
p_holm <- p.adjust(pvals, method = "holm")
p_bonf  <- p.adjust(pvals, method = "bonferroni")
# Combine for reporting
results <- data.frame(
  Effect = rownames(anova_res)[rows_biling],
  `LR Chisq` = anova_res[rows_biling, "LR Chisq"],
  df = anova_res[rows_biling, "Df"],
  p_raw = round(pvals, 5),
  p_holm = round(p_holm, 5),
  p_bonf = round(p_bonf, 5)
)
results
exp(cbind(OR = coef(model_stable), confint(model_stable)))
effectsize(model_stable)
summary(model_stable)
eta_squared(model_stable) 
```

We fit a regression model with StableCU as the dependent variable. Predictors included A0, Bilingualism; interaction terms captured moderation by age, gender, and ethnicity where specified.

### 2.2 Cognitive Abilities as Outcome

### 2.2.1 EFA for cognitive measures 

```{r}
# Reverse Trails A and B so higher = better
HABS_HD_longitudinal$Trails_A_Reversed <- -scale(HABS_HD_longitudinal$Trails_A_Time)
HABS_HD_longitudinal$Trails_B_Reversed <- -scale(HABS_HD_longitudinal$Trails_B_Time)

### Select the target cognitive measures  
HABS_data_z2 = subset(HABS_HD_longitudinal, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total','Med_ID','Visit_ID'))
HABS_data_z2$Med_ID<-as.factor(HABS_data_z2$Med_ID)
HABS_data_z2$Visit_ID<-as.factor(HABS_data_z2$Visit_ID)
HABS_data_z2 <- HABS_data_z2  # make a copy
HABS_data_z2[sapply(HABS_data_z2, is.numeric)] <- scale(HABS_data_z2[sapply(HABS_data_z2, is.numeric)])

## create a z-scored dataset for cognitive measures
efa_data2 = subset(HABS_data_z2, select=c('Trails_A_Reversed', 'Trails_B_Reversed', 'Digit_Symbol_Substitution', 'DSF_Total', 'DSB_Total', 'LM1_AB_Total', 'LM2_AB_Total', 'SEVLT_T1235_Total', 'SEVLT_DR_Total', 'FAS_Total', 'Animal_Total')) 
```

```{r}
psych::fa.parallel(efa_data2, fa = "pc")
#### Parallel analysis suggests that the number of components =  2 

# Run PCA
pca_result <- psych::principal(efa_data2, nfactors = 2, rotate = "oblimin")
# View results
print(pca_result, digits = 2, cutoff = 0.3)
pca_result$Vaccounted
```

```{r}
two_factor_efa_fit2 = lavaan::efa(data =efa_data2, nfactors=2, rotation='oblimin', missing="fiml")
summary(two_factor_efa_fit2, fit.measures = TRUE, standardized = TRUE)

predicted_scores2 <- lavPredict(two_factor_efa_fit2)
predicted_scores2 <- as.data.frame(predicted_scores2)
colnames(predicted_scores2) <- c("Executive_Function", "Episodic_Memory")

HABS_HD_longitudinal <- cbind(HABS_HD_longitudinal, predicted_scores2)
```
### 2.2.2 Analysis on Cognitive Performance: Executive_Function domain 

```{r}
HABS_HD_longitudinal$CatPractice <- as.factor(HABS_HD_longitudinal$Visit_ID)
a1<-lmer(Executive_Function ~  A0 * dA * Bilingualism * Gender * Ethnicity + EducationYears + CatPractice + (1|Med_ID), data = HABS_HD_longitudinal)
car::Anova(a1)
anova_res <- car::Anova(a1, type = 2)
anova_res <- as.data.frame(anova_res) %>%
  mutate(Effect = rownames(anova_res))
biling_terms <- anova_res %>%
  filter(grepl("dA:Bilingualism", Effect))

pvals <- biling_terms$`Pr(>Chisq)`

biling_terms <- biling_terms %>%
  mutate(
    p_holm = p.adjust(pvals, method = "holm"),
    p_bonf = p.adjust(pvals, method = "bonferroni") )

biling_terms <- biling_terms %>%
  mutate(across(c(Chisq, `Pr(>Chisq)`, p_holm, p_bonf), ~round(.,2)))

results <- biling_terms %>%
  dplyr::select(Effect, Chisq, df = Df, p_raw = `Pr(>Chisq)`, p_holm, p_bonf)

results
eta_squared(a1)
test(emtrends(a1, ~ Bilingualism * Ethnicity, var = "dA"))
effectsize(a1)
```

### 2.2.2 Analysis on Cognitive Performance: Long-term Memory (Episodic_Memory)
```{r}
a2<-lmer(Episodic_Memory ~  A0 * dA * Bilingualism  * Gender  * Ethnicity + EducationYears + CatPractice + (1|Med_ID), data = HABS_HD_longitudinal)
car::Anova(a2)
# Extract p-values for fixed effects
anova_res <- car::Anova(a2, type = 2)
anova_res <- as.data.frame(anova_res) %>%
  mutate(Effect = rownames(anova_res))
biling_terms <- anova_res %>%
  filter(grepl("dA:Bilingualism", Effect))
pvals <- biling_terms$`Pr(>Chisq)`
biling_terms <- biling_terms %>%
  mutate(
    p_holm = p.adjust(pvals, method = "holm"),
    p_bonf = p.adjust(pvals, method = "bonferroni") )

biling_terms <- biling_terms %>%
  mutate(across(c(Chisq, `Pr(>Chisq)`, p_holm, p_bonf), ~round(.,3)))

results <- biling_terms %>%
  dplyr::select(Effect, Chisq, df = Df, p_raw = `Pr(>Chisq)`, p_holm, p_bonf)

results
test(emtrends(a2, ~ Bilingualism , var = "dA"))
car::Anova(a2) 
eta_squared(a2)
effectsize(a2)

```

dA:Bilingualism is significant, indicating bilingual protective effect


Plot for Episodic_Memory Performance
```{r}
plot_spaghetti_collapsed <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, plot.title, plot.subtitle = NULL) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    # raw points
    geom_point(alpha = 0.3, size = 2, color = "darkorange") +
    # raw trajectories per participant
    geom_line(aes(group = Med_ID), linetype = "dotted", alpha = 0.2, color = "#377EB8") +
    # smooth trend line (quadratic)
    geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE, color = "#377EB8", size = 1) +
    labs(
      x = x_label,
      y = y_label,
      title = plot.title,
      subtitle = plot.subtitle
    ) +
    ylim(ymin, ymax) +
    theme_minimal() +
     theme_bw() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
      plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
      legend.position = "top",
      legend.title = element_text(size = 14, family = "serif"),
      legend.text = element_text(size = 14, family = "serif"),
      axis.title.x = element_text(size = 14, family = "serif"),
      axis.title.y = element_text(size = 14, family = "serif"),
      axis.text.x = element_text(size = 13, family = "serif"),  
      axis.text.y = element_text(size = 13, family = "serif"),
      strip.text = element_text(size = 14, family = "serif")
    )
}

p61 <- plot_spaghetti_collapsed(
  data = HABS_HD_longitudinal,
  x_var = "Age",
  y_var = "Episodic_Memory",
  x_label = "Age",
  y_label = "Episodic Memory",
  ymin = min(HABS_HD_longitudinal$Episodic_Memory, na.rm = TRUE),
  ymax = max(HABS_HD_longitudinal$Episodic_Memory, na.rm = TRUE),
  plot.title = "Episodic Memory Changes Across Age",
  plot.subtitle = "Dotted lines: individual trajectories; Blue line: quadratic trend"
)
p61

```


Bilingualism was associated with a slower decline in episodic memory over time.
```{r}
# 1. Fit the model
m1 <- lmer(Episodic_Memory ~ A0 * dA * Ethnicity * Gender * Bilingualism +
             CatPractice + EducationYears + (1|Med_ID),
           data = HABS_HD_longitudinal)

# 2. Filter raw data for dA <= 6
raw_data_filtered <- HABS_HD_longitudinal %>% 
  filter(dA <= 6)

# 3. Get adjusted predictions for dA × Bilingualism, limiting dA to <= 6
pred <- emmeans(m1, ~ dA * Bilingualism,
                at = list(dA = seq(min(raw_data_filtered$dA),
                                   max(raw_data_filtered$dA),
                                   length.out = 50)))
pred_df <- as.data.frame(pred)

# 4. Plot
p62<-ggplot() +
  # Raw participant trajectories
  geom_line(data = raw_data_filtered,
            aes(x = dA, y = Episodic_Memory, group = Med_ID),
            color = "orange", linetype = "dotted", alpha = 0.5) +
  geom_point(data = raw_data_filtered,
             aes(x = dA, y = Episodic_Memory),
             color = "orange", alpha = 0.5, size = 1.5) +
  # Model-predicted lines
  geom_line(data = pred_df,
            aes(x = dA, y = emmean, color = Bilingualism),
            size = 1.2) +
  # Confidence ribbons
  geom_ribbon(data = pred_df,
              aes(x = dA, ymin = lower.CL, ymax = upper.CL, fill = Bilingualism),
              alpha = 0.2, inherit.aes = FALSE) +
  # Labels
  labs(
    x = "Time Differences between Visits (dA)",
    y = "Episodic Memory (adjusted)",
    title = "Episodic Memory Trajectories by Bilingualism",
    subtitle = "Orange dots: individual raw data; Solid lines: model predictions"
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
   theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16, family = "serif"),
    plot.subtitle = element_text(face = "plain", hjust = 0.5, size = 14, family = "serif"),
    legend.position = "top",
    legend.title = element_text(size = 14, family = "serif"),
    legend.text = element_text(size = 14, family = "serif"),
    axis.title.x = element_text(size = 14, family = "serif"),
    axis.title.y = element_text(size = 14, family = "serif"),
    axis.text.x = element_text(size = 13, family = "serif"),  
    axis.text.y = element_text(size = 13, family = "serif"),
    strip.text = element_text(size = 14, family = "serif")
  )
p62
combined_plot <- ggarrange(
  p61, p62,
  ncol = 1, nrow = 2,
  labels = c("A", "B"),      # optional panel labels
  #common.legend = False,      
  legend = "top"              # legend position
)

ggsave(
  "Episodic_Memory_longitudinal_combined.png",
  plot = combined_plot,
  width = 11,
  height = 16,
  dpi = 800
)

ggsave("Episodic_Memory_longitudinal2.png", plot = p61, 
       width = 10, height = 6, units = "in", 
       dpi = 800) 
```







